
---
title: "VBB GTFS data analysis"
output:
  flexdashboard::flex_dashboard:
  orientation: rows # rows
  vertical_layout: scroll # fill
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(sf)
library(tidytransit)
library(tmap)
library(readxl)
library(flexdashboard)
library(sfnetworks)
library(tidygraph)
library(fuzzyjoin)

vbb_gtfs <- read_gtfs("BVG_VBB_bereichsscharf.zip")

vbb_shapes <- shapes_as_sf(vbb_gtfs$shapes)
vbb_stops <- stops_as_sf(vbb_gtfs$stops)


berlin_bezirke <- st_read("shapes/GRENZEN_1163761-6656443_berlinbezirke.shp")

ew_be <- read_xlsx("ew_berlin.xlsx") %>% 
  select("gemeinde", "einwohner", "dichte")

ew_bb_be <- read_xlsx("GemVerz.xlsx") %>% 
  select("gemeinde", "einwohner", "flaeche") %>%
  filter(!is.na(flaeche)) %>% 
  mutate(dichte = einwohner/flaeche*100) %>% 
  full_join(ew_be)

gemeinde_shapes <- st_read("shapes/GRENZEN_1163761-6656443_gemeinden.shp") %>%
  union(berlin_bezirke) %>% 
  st_as_sf() %>% 
  left_join(ew_bb_be, by = c("NAME" = "gemeinde"))
  
frequencies <- get_route_frequency(vbb_gtfs)

stops_departures <- vbb_gtfs$stop_times %>% 
  group_by(stop_id) %>% 
  summarise(anzahl_abfahrten = n()) %>% 
  left_join(vbb_stops, by = "stop_id") %>% 
  st_as_sf()

bb_eisenbahn <- st_read("shapes/gis_osm_railways_free_1.shp")
```

Departures <> Population Density
==================

How many trips depart from a stop in relation to the population density?

```{r include=FALSE}
# 060135001111 // 450009182002
around_friedrichshagen <- vbb_stops %>% 
  filter(stop_id == "060135001111") %>% 
  st_as_sf(crs=4326) %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_buffer(15000)

stops_around_friedrichshagen <- stops_departures %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_friedrichshagen)

gemeinden_around_friedrichshagen <- gemeinde_shapes %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_friedrichshagen)
```


```{r}
tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(around_friedrichshagen) +
  tm_borders() +
  tm_shape(gemeinden_around_friedrichshagen) +
  tm_polygons(col = "dichte", title="population density", alpha = 0.4, palette = '-viridis', id = "NAME", style = "log10",  popup.vars = c("Gemeindename" = "NAME", "Einwohnerdichte" = "dichte")) +
  tm_shape(stops_around_friedrichshagen) +
  tm_dots(size = 0.2, col = "anzahl_abfahrten", title = "number of departures per day", palette = '-viridis', style = "log10", id = "stop_name")
```



Availability of Stops
==================

Are there any areas in Potsdam where there is no stop in 400 meters of walking?

"Nahverkehrsplan" for city of Potsdam defines 300-500 meters walking distance as criteria for good service quality. Map shows circles around stops with a radius of 400 meters, regardless how often services run at the stops.

```{r include=FALSE}

potsdam_stops <- vbb_stops %>% 
  filter(str_detect(stop_name, "Potsdam(,| )")) %>% 
  st_as_sf(crs=4326) %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_buffer(400)

potsdam_shape <- gemeinde_shapes %>% 
  filter(NAME == "Potsdam")
```


```{r}
tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(potsdam_shape) +
  tm_borders(col = 'red', lwd = 2) +
  tm_shape(potsdam_stops) +
  tm_polygons(id = "stop_name", alpha = 0.3, interactive = FALSE, border.alpha = 0.2)
```

Isochrone Map
==================

How far can you go in the VBB area in the time of our lecture (departing later than 4:15 PM and arriving earlier than 5:45 PM) starting at Berlin Hauptbahnhof station?

```{r}
travel_times_BL_0 <- filter_stop_times(vbb_gtfs, "2021-02-15", "16:15:00", "17:45:00") %>% 
  travel_times(c("S+U Berlin Hauptbahnhof", "S+U Berlin Hauptbahnhof (tief)"), max_transfers = 0, return_coords = TRUE) %>% 
  mutate(travel_time_hms = hms::as_hms(travel_time))



travel_times_BL_5 <- filter_stop_times(vbb_gtfs, "2021-02-15", "16:15:00", "17:45:00") %>% 
  travel_times(c("S+U Berlin Hauptbahnhof", "S+U Berlin Hauptbahnhof (tief)"), max_transfers = 5, return_coords = TRUE) %>% 
  mutate(travel_time_hms = hms::as_hms(travel_time))

```

# of Stops reachable without changes
-------------------------------------

### # of Stops reachable without changes

```{r}
valueBox(nrow(travel_times_BL_0), icon = "fa-bus")
```


### 0 transfers

```{r}
tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(bb_eisenbahn)+
  tm_lines(col="fclass", legend.col.show = FALSE, scale =3, interactive = FALSE)+
  tm_shape(st_as_sf(travel_times_BL_0, coords = c("to_stop_lon", "to_stop_lat"))) +
  tm_dots(col = "transfers", style = "cat", palette = '-viridis', id = "to_stop_name", popup.vars = c("travel_time_hms", "transfers"), title = "Number of transfers", legend.show = FALSE)
  
```


# of Stops reachable up to five changes
-------------------------------------

### # of Stops reachable with up to five changes

```{r}
valueBox(nrow(travel_times_BL_5), icon = "fa-bus")
```


### 5 transfers

```{r}
tmap_mode("view")
tm_basemap("OpenStreetMap.Mapnik") +
    tm_shape(bb_eisenbahn)+
  tm_lines(col="fclass", legend.col.show = FALSE, scale=3, interactive=FALSE)+
  tm_shape(st_as_sf(travel_times_BL_5, coords = c("to_stop_lon", "to_stop_lat"))) +
  tm_dots(col = "transfers", style = "cat", palette = '-viridis', id = "to_stop_name", popup.vars = c("travel_time_hms", "transfers"), title = "Number of transfers")

```


Suburb accessibility 
==================

Don't want to spend all your money on rent? A look at service frequencies on transit lines around the borders of Berlin to determine where a comfortable car-free life might be possible.

```{r}
vbb_gtfs <- set_servicepattern(vbb_gtfs)
# buch actually 060135001111, but many many route artifacts

around_buch <- vbb_stops %>% 
  filter(stop_id == "450009182002") %>% 
  st_as_sf(crs=4326) %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_buffer(15000)

stops_around_buch <- stops_departures %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_buch)

gemeinden_around_buch <- gemeinde_shapes %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_buch)

# idee: wochenende und nachts
dichte_tags체ber <- get_route_frequency(vbb_gtfs, start_hour = 6, end_hour = 10, dow=c(1,1,1,1,1,0,0)) # weekdays
dichte_woche_sp채tabend <- get_route_frequency(vbb_gtfs, start_hour = 21, end_hour = 24, dow=c(1,1,1,1,1,0,0))
dichte_wochenende_party <- get_route_frequency(vbb_gtfs, start_hour = 0, end_hour = 4, dow=c(0,0,0,0,0,1,1))

vbb_sf <- gtfs_as_sf(vbb_gtfs)
vbb_linien_sf <- get_route_geometry(vbb_sf) 

vbb_linien_sf_tag <- vbb_linien_sf %>% 
  inner_join(dichte_tags체ber, by ="route_id") 

vbb_linien_sf_spaet<- vbb_linien_sf %>% inner_join(dichte_woche_sp채tabend, by="route_id",suffix=c(".tag",".spaet"))
  
vbb_linien_sf_nachts <- vbb_linien_sf %>% inner_join(dichte_wochenende_party, by="route_id",suffix=c("x",".party"))
```

Weekdays, Rush Hour (0600-1000)
-------------------------------------

### Weekdays, Rush Hour (0600-1000)

```{r}
lines_around_buch <- vbb_linien_sf_tag %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_buch)

tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(around_buch) +
  tm_borders() +
  tm_shape(lines_around_buch) +
  tm_lines(col="median_headways", palette = '-viridis', title.col = "median route headwys")
  
#style="fixed", breaks=c(0,15,30,60,120,240) --> will nicht richtig
```

Weekend, early morning (0000-0400)
-------------------------------------

### Weekend, early morning (0000-0400)

(post-night-life-rush-hour)

```{r}
lines_around_buch_nachts <- vbb_linien_sf_spaet %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_intersection(around_buch)

tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(around_buch) +
  tm_borders() +
  tm_shape(lines_around_buch_nachts) +
  tm_lines(col="median_headways", style = "fixed", palette = '-viridis',breaks=c(0,15,30,60,120,240), title.col="median route headways")
```

# Stops per X
Numbers are skewed inside dense areas because of the way VBB defines stops.

```{r preparation}
# buch actually 060135001111, but many many route artifacts

around_buch <- vbb_stops %>% 
  filter(stop_id == "450009182002") %>% 
  st_as_sf(crs=4326) %>% 
  st_transform("+proj=utm +zone=32") %>% 
  st_buffer(15000)

my_stops <- vbb_stops %>% 
  st_transform(crs=3857)

stopfeats <- vbb_stops %>% 
  filter(stop_id=="450009182002") %>%  
  st_transform("+proj=utm +zone=32") %>% 
  st_intersects(around_buch)

abc <- gemeinde_shapes %>% mutate(stops = st_intersects(geometry, my_stops))
def <- abc %>% mutate(stopcount = lengths(stops)) %>% mutate(persons_per_stop = einwohner/stopcount, stops_per_area=stopcount/flaeche, stops_per_density=stopcount/dichte)

gemeindestopdensity = def %>% select(NAME,persons_per_stop,stops_per_area,stops_per_density,geometry) %>% filter()
```

### How many stops exist per person per square kilometer?
(Stops per Area would more or less be a population density plot)

```{r}
tm_basemap("OpenStreetMap.Mapnik") +
  tm_shape(bb_eisenbahn)+
  tm_lines(scale=2)+
  tm_shape(gemeindestopdensity) +
  tm_polygons(col="stops_per_density", alpha=0.6, palette = '-viridis', style = "pretty", id = "NAME", title="Stops per inhabitant per sq km")
```


Sources
==================

Sources:

- [VBB GTFS dataset](https://daten.berlin.de/datensaetze/vbb-fahrplandaten-gtfs)
- [Shapefiles Berlin](https://geobroker.geobasis-bb.de/gbss.php?MODE=GetProductInformation&PRODUCTID=00fdc3fb-3bc1-4548-bca2-e735fb11c974)
- [Shapefiles Brandenburg](https://geobroker.geobasis-bb.de/gbss.php?MODE=GetProductInformation&PRODUCTID=00fdc3fb-3bc1-4548-bca2-e735fb11c974)
- [Population and Area of municipalities in Brandenburg](https://www.statistik-berlin-brandenburg.de/statistiken/statistik_SB.asp?Ptyp=700&Sageb=12015&creg=BBB&anzwer=6)
- [Shapefiles of Brandenburg+Berlin Railway lines](https://download.geofabrik.de/europe/germany/brandenburg-latest-free.shp.zip)
- [Population density of Berlin Bezirke](https://www.statistik-berlin-brandenburg.de/regionalstatistiken/r-gesamt_neu.asp?Ptyp=410&Sageb=12015&creg=BBB&anzwer=6) 

